Discriminative Sensing Based on Signal Processing for Information Security Analysis

### 2017-10-27 docs/thesis/chapters/2_eigen_similarity/eigen_similarity.tex
introducion...
conclusion and introduction

\chapter{Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks}
\label{ch:evaluation}

@@

In the context of anomaly-based schemes, this work proposes a statistical approach based on signal processing techniques for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 dataset \citep{osanaiye2016distributed}, which is a well known network traffic dataset. Furthermore, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

@@

This chapter is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model and the evaluated datasets. Section \ref{sec:prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

\section{Related Works}
\label{sec:relatedworks}

@@

\section{Data Model}
\label{sec:datamodel}

@@

This section presents details of the experimental scenario and the selected cases of the DARPA dataset, along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:DarpaDataset} discusses the use of the DARPA dataset for evaluation of the proposed approach.

\subsection{Analyzed Scenario and Data Collection}
\label{sec:DataCollection}

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, flood and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=6cm, width=9cm]{figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:fig1}
\end{figure}

@@

\subsection{Modeling Data}
\label{sec:ModelingData}

@@

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:fig3} depicts an example of the legitimate traffic obtained during experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:fig3}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:fig4}
\end{figure}

@@

\subsection{Synflood, Fraggle and Port scan}
\label{sec:SynfloodFraggleandPortscan}

@@

With respect to the synflood attacks, the attacker sends a large quantity and concurrent successive SYN requests to a target, in order to consume resources and cause a DoS. Figure \ref{fig:fig5} depicts an example of a synflood attack carried out in a real computer network. In an interval of ten minutes, more than 210,000 packets are sent as a synflood attack. This network traffic behavior can be considered an abnormal behavior of network traffic, especially since it is concentrated in a short period of time and presents similar outstanding traffic during the time under attack.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:fig5}
\end{figure}

With respect to the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network.

\begin{figure}[h!]
     \centering 
     \includegraphics[height=4.5cm, width=9cm]{figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:fig6}
\end{figure}

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[height=5cm, width=9cm]{figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:fig7}
\end{figure}

Figure \ref{fig:fig7} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

\subsection{The DARPA Dataset}
\label{sec:DarpaDataset}

@@

The data modeling follows the method described by the Subsection \ref{sec:ModelingData}, with time frames of 20 minutes, packet counting aggregation by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

\section{Proposed Framework for Detection and Identification of Network Attacks}
\label{sec:prop_getv}

This section describes the proposed technique to detect synflood, fraggle and port scan, according to Figure \ref{fig:fig80}, which represents the overview of the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvelue for each $q$-th time frame. Next, in Subsection \ref{sec:prop_MOSSchemes}, we show how to apply the eiganvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[height=11cm, width=9cm]{figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:fig80}
\end{figure}

\subsection{Largest Eigenvalue by Time Frames}
\label{sec:prop_LargestEigenvaluebyTimeFrames}

@@

According to flood and port scan attacks' behavior, flood attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of flood attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

@@

\subsection{MOS Schemes}
\label{sec:prop_MOSSchemes}

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to \ref{sec:mos}.

In our previous work \cite{tenorio2013greatest}, the accuracy of AIC, MDL, EDC, RADOI, EFT and SURE schemes are evaluated for synflood and port scan attack detection, showing that EDC and EFT are effective for detecting this kind of attacks. The present work extends that evaluation to also analyze the effectiveness of the listed MOS schemes for fraggle attack detection, as shown in Section \ref{sec:experimentalresults}.

\subsection{Eigenvalue Analysis}
\label{sec:prop_EigenvalueAnalysis}

@@

In Subsection \ref{sec:prop_MOSSchemes}, we obtained only if $\hat{d} = 1$ or $\hat{d} = 0$. However, if $\hat{d} = 1$, the MOS schemes do not provide any information about the $q$-th time frame under attack. The identification of the $q$-th time frame under attack can be carried out through a eigenvalues analysis.

The largest eigenvalue analysis for estimating the $q$-th time frames that are under attack can be expressed according to Algorithm \ref{alg:alg01}, where $\rm{\boldsymbol{\hat{q}}}_{\rm max} \in \mathbb{R}^{\hat{d}}$ denotes a vector of the $q$-th time frames under attack, which is the $q$-th indexes corresponding to the $\hat{d}$ largest eigenvalues of $\boldsymbol{e}_{\rm max}$. Algorithm \ref{alg:alg01} initially identifies the largest value of $\boldsymbol{e}_{\rm max}$, according to Line 2 of Algorithm \ref{alg:alg01}, and its correspondent index, according to Lines 4 and 5 of Algorithm \ref{alg:alg01}. Subsequently, the largest value is removed of $\boldsymbol{e}_{\rm max}$, according to Line 8 of Algorithm \ref{alg:alg01}, and a new iteration is performed until $\boldsymbol{e}_{\rm max} = \{\}$.

\begin{algorithm}[h!]
	% \caption{Detection of Time Frames Under Attack}
 %  	\label{alg:alg01}
	% \begin{algorithmic}[1]
	% 	\show\LOOP
	%     \ALOOP {$f = 1$ until $f == \hat{d}$} 
	%     	    \STATE $q_{\rm value} = \argmax\limits_{\lambda}  \hspace{1 mm} \boldsymbol{e}_{\rm max}$
 %    	    	    \ALOOP {$i = 1$ until $i == Q$} 
	% 			\IF {$\boldsymbol{e}_{\rm max}^{(i)} == q_{\rm value}$}
	% 			    \STATE $\boldsymbol{\hat{q}}_{\rm max}^{(f)} = i$
	% 			\ENDIF
 %        		\ENDALOOP
	%     		\STATE $\boldsymbol{e}_{\rm max} \rightarrow \boldsymbol{\hat{q}}_{\rm max}^{(f)}$
 %   		\ENDALOOP
	% \end{algorithmic}
\end{algorithm}

@@

\subsection{Eigen Similarity Analysis}
\label{sec:prop_EigenSimilarityAnalysis}

@@

This subsection describes the proposed eigen similarity analysis for detailed attack identification, in complement to the attack estimation carried out through MOS schemes and eigenvalue analysis. In Subsection \ref{sec:prop_TimeSimilarityAnalysis} we present the eigen similarity analysis for identification of time under attack. Next, in Subsection \ref{sec:prop_PortSimilarityAnalysis}, we show how to apply the eigen similarity analysis in order to identify network ports under attack.

\subsubsection{Time Similarity Analysis}
\label{sec:prop_TimeSimilarityAnalysis}

@@

Figure \ref{fig:fig8} ilustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\boldsymbol{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected.

\begin{figure}[h!]
     \includegraphics[height=4cm, width=12.4cm]{figures/incremental.eps}
     \caption{Traffic selection for incremental approach.}
     \label{fig:fig8}
\end{figure}

@@

The eigen similarity analysis can also be applied in an individual fashion, where each $n$-th minute must be individually appended into $\boldsymbol{X}^{(q)}$, as shown by Figure \ref{fig:fig9}.

\begin{figure}[h!]
     \includegraphics[height=4cm, width=12.4cm]{figures/individualized.eps}
     \caption{Traffic selection for individual approach.}
     \label{fig:fig9}
\end{figure}

The incremental and the individual approaches can be combined to obtain the incremental individualized approach, where each minute is incrementally appended into the selected $\boldsymbol{X}^{(q)}$ for obtaining $\boldsymbol{v}_{(n)}$ to similarity analysis of the $n$-th minute, until detect the first $n$-th minute under attack. Subsequently, $\boldsymbol{X}_n$ becomes the new reference of traffic without network attack and each subsequent minute must have its similarity individually evaluated, as shown in Figure \ref{fig:fig2}.

\begin{figure}[h!]
     \includegraphics[height=4cm, width=12.4cm]{figures/incremental_individualized.eps}
     \caption{Traffic selection for incremental individualized approach.}
     \label{fig:fig2}
\end{figure}

This approach of incremental similarity analysis followed by individual analysis after an attack detection allows to identify the attack period, highlighting the first and last time under attack. This identification is possible due to the variation of the most significant eigenvectors, which becomes more significant when compared a traffic under attack against a traffic with no attack, according to results which are discussed in Section \ref{sec:experimentalresults}.

\subsubsection{Port Similarity Analysis}
\label{sec:prop_PortSimilarityAnalysis}

@@

\section{Experiments and Results}
\label{sec:experimentalresults}

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the experimental scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames for the experimental scenario. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection in the simulated dataset. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed flood and port scan identification for the experimental scenario. Section \ref{sec:DarpaEvaluation} presents the results of the largest eigenvalue analysis, model order selection and the eigenvalue analysis for flood and probe attack detection in the DARPA 1998 dataset.

\subsection{Experimental Scenario}
\label{sec:AnalyzedScenario}

@@

\subsection{Largest Eigenvalues Analysis}
\label{sec:largesteigenvaluesanalysis}

For the evaluation of MOS Schemes accuracy for flood and port scan detection, the framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:prop_getv}. Through eigenvalue analysis of traffic with flood or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

\begin{figure}[h!]
	\centering
     \includegraphics[height=6cm, width=8cm]{figures/eigenvalues_synflood.eps} 
     \caption{Eigenvalues of the sample covariance matrix (synflood).}
     \label{fig:fig10}
\end{figure}

Figure \ref{fig:fig10} depicts the eigenvalues calculated from sample covariance matrix of the network traffic used to evaluate the synflood attack identification. In Figure \ref{fig:fig10}, the largest eigenvalue related to the simulated synflood attack ($q = 4$) stands out significantly from the other eigenvalues.

Figure \ref{fig:fig11} illustrates the eigenvalues calculated from sample covariance matrix of the matrix used for fraggle attack detection. In Figure \ref{fig:fig10}, the largest eigenvalue related to the simulated synflood attack ($q = 5$) stands out significantly from the other eigenvalues, in accordance with the result shown in Figure \ref{fig:fig10} for the synflood attack analysis.

\begin{figure}[h!]
	\centering
     \includegraphics[height=6cm, width=8cm]{figures/eigenvalues_fraggle.eps}
     \caption{Eigenvalues of the sample covariance matrix (fraggle).}
     \label{fig:fig11}
\end{figure}

Figure \ref{fig:fig12} depicts the eigenvalues calculated from covariance matrix of zero mean and unitary standard deviation variables, of the network traffic matrix evaluated for port scan detection.

>>>>

\chapter{Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks}
\label{ch:2_mos_eig_sim}

@@

In the context of anomaly-based schemes, this thesis proposes a statistical approach based on signal processing techniques for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this thesis models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy and performance of the proposed framework applied to a experimental scenario and to the DARPA 1998 dataset \citep{osanaiye2016distributed}, which is a well known network traffic dataset. Furthermore, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

@@

This chapter is organized as follows. In Section \ref{sec:2_relatedworks}, related works are discussed. Section \ref{sec:2_datamodel} presents the data model and the evaluated datasets. Section \ref{sec:2_prop_getv} describes the proposed framework for blind and automatic detection of flood and probe attacks. Section \ref{sec:2_experimentalresults} discusses the experimental validation and presents the results, and Section \ref{sec:2_Complexity} discusses the computational complexity of the proposed framework and evaluates the required processing time for tested scenarios. Section \ref{sec:2_conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{apx:a_mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

\section{Related Works}
\label{sec:2_relatedworks}

@@

\section{Data Model}
\label{sec:2_datamodel}

@@

This section presents details of the experimental scenario and the selected cases of the DARPA dataset, along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:2_DataCollection} describes the environment and scenario adopted in order to reproduce flood and probe attacks. Subsection \ref{sec:2_ModelingData} presents how network traffic can be modeled as signal superposition. Subsection \ref{sec:2_SynfloodFraggleandPortscan} details the traffic of synflood, fraggle and port scan attacks, and Subsection \ref{sec:2_DarpaDataset} discusses the use of the DARPA dataset for evaluation of the proposed approach.

\subsection{Analyzed Scenario and Data Collection}
\label{sec:2_DataCollection}

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, flood and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:2_fig1}.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig09.png}
     \caption{Scenario to reproduce legitimate traffic, noise, flood and port scan.}
     \label{fig:2_fig1}
\end{figure}

@@

\subsection{Modeling Data}
\label{sec:2_ModelingData}

@@

The legitimate traffic $\boldsymbol{U}^{(q)}$ is characterized by the traffic from user's operations. When a user accesses a web page, for example, there is the corresponding TCP/IP traffic to request the page, as well as there is the traffic required to domain name resolution. Figure \ref{fig:2_fig3} depicts an example of the legitimate traffic obtained during experiments.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig03.png}
     \caption{Traffic from user's operations, that can be characterized by web access, traffic of well-known applications or network protocols.}
     \label{fig:2_fig3}
\end{figure}

The traffic that is not associated with user's operations and with malicious traffic is modeled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:2_fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig04.png}
     \caption{Network traffic of user independent operations for network management.}
     \label{fig:2_fig4}
\end{figure}

@@

\subsection{Synflood, Fraggle and Port scan}
\label{sec:2_SynfloodFraggleandPortscan}

@@

With respect to the synflood attacks, the attacker sends a large quantity and concurrent successive SYN requests to a target, in order to consume resources and cause a DoS. Figure \ref{fig:2_fig5} depicts an example of a synflood attack carried out in a real computer network. In an interval of ten minutes, more than 210,000 packets are sent as a synflood attack. This network traffic behavior can be considered an abnormal behavior of network traffic, especially since it is concentrated in a short period of time and presents similar outstanding traffic during the time under attack.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig05.png}
     \caption{A large quantity of SYN requests to a target, in order to cause a DoS.}
     \label{fig:2_fig5}
\end{figure}

With respect to the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:2_fig6} depicts an example of the fraggle attack in a real computer network.

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig06.png}
     \caption{Large amount of “UDP echo” requests and replies, causing packet flooding.}
     \label{fig:2_fig6}
\end{figure}

@@

\begin{figure}[h!]
     \centering 
     \includegraphics[width=11cm]{figures/fig07.png}
     \caption{Connection attempts in order to identify active ports.}
     \label{fig:2_fig7}
\end{figure}

Figure \ref{fig:2_fig7} depicts an example of the port scan attack in a real computer network. Note that the traffic is composed of two packets for each TCP port and one UDP packet to each port. The incoming and outgoing packets analysis, for each port, shows the high correlation and similarity of TCP and UDP traffic during the simulated port scan attack.

\subsection{The DARPA Dataset}
\label{sec:2_DarpaDataset}

@@

The data modeling follows the method described by the Subsection \ref{sec:2_ModelingData}, with time frames of 20 minutes, packet counting aggregation by minute and considering the traffic to the following ports: 20, 21, 22, 23, 25, 79, 80, 88, 107, 109, 110, 113, 115, 143, 161, 389, 443.

\section{Proposed Framework for Detection and Identification of Network Attacks}
\label{sec:2_prop_getv}

This section describes the proposed technique to detect synflood, fraggle and port scan, according to Figure \ref{fig:2_fig80}, which represents the overview of the proposed framework for detection and identification of network attacks. In Subsection \ref{sec:2_prop_LargestEigenvaluebyTimeFrames} we present the steps for extraction of the largest eigenvelue for each $q$-th time frame. Next, in Subsection \ref{sec:2_prop_MOSSchemes}, we show how to apply the eiganvalues on the MOS scheme in order to detect the attack. In Subsection \ref{sec:2_prop_EigenvalueAnalysis}, we present the eigenvalue analysis to identify the time frames detected as under attack, and the Subsection \ref{sec:2_prop_EigenSimilarityAnalysis} describes the similarity analysis evaluated for detailed attack identification.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/mos_eigen_similarity.eps}
     \caption{Overview of The Framework for Detection and Identification of Network Attacks.}
     \label{fig:2_fig80}
\end{figure}

\subsection{Largest Eigenvalue by Time Frames}
\label{sec:2_prop_LargestEigenvaluebyTimeFrames}

@@

According to flood and port scan attacks' behavior, flood attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:2_experimentalresults}, which shows that the main components of flood attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

@@

\subsection{MOS Schemes}
\label{sec:2_prop_MOSSchemes}

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to \ref{apx:a_mos}.

In our previous work \cite{tenorio2013greatest}, the accuracy of AIC, MDL, EDC, RADOI, EFT and SURE schemes are evaluated for synflood and port scan attack detection, showing that EDC and EFT are effective for detecting this kind of attacks. The present work extends that evaluation to also analyze the effectiveness of the listed MOS schemes for fraggle attack detection, as shown in Section \ref{sec:2_experimentalresults}.

\subsection{Eigenvalue Analysis}
\label{sec:2_prop_EigenvalueAnalysis}

@@

In Subsection \ref{sec:2_prop_MOSSchemes}, we obtained only if $\hat{d} = 1$ or $\hat{d} = 0$. However, if $\hat{d} = 1$, the MOS schemes do not provide any information about the $q$-th time frame under attack. The identification of the $q$-th time frame under attack can be carried out through a eigenvalues analysis.

The largest eigenvalue analysis for estimating the $q$-th time frames that are under attack can be expressed according to Algorithm \ref{alg:2_alg01}, where $\rm{\boldsymbol{\hat{q}}}_{\rm max} \in \mathbb{R}^{\hat{d}}$ denotes a vector of the $q$-th time frames under attack, which is the $q$-th indexes corresponding to the $\hat{d}$ largest eigenvalues of $\boldsymbol{e}_{\rm max}$. Algorithm \ref{alg:2_alg01} initially identifies the largest value of $\boldsymbol{e}_{\rm max}$, according to Line 2 of Algorithm \ref{alg:2_alg01}, and its correspondent index, according to Lines 4 and 5 of Algorithm \ref{alg:2_alg01}. Subsequently, the largest value is removed of $\boldsymbol{e}_{\rm max}$, according to Line 8 of Algorithm \ref{alg:2_alg01}, and a new iteration is performed until $\boldsymbol{e}_{\rm max} = \{\}$.

\begin{algorithm}[h!]
	\caption{Detection of Time Frames Under Attack}
  	\label{alg:2_alg01}
	% \begin{algorithmic}[1]
	% 	\show\LOOP
	%     \ALOOP {$f = 1$ until $f == \hat{d}$} 
	%     	    \STATE $q_{\rm value} = \argmax\limits_{\lambda}  \hspace{1 mm} \boldsymbol{e}_{\rm max}$
 %    	    	    \ALOOP {$i = 1$ until $i == Q$} 
	% 			\IF {$\boldsymbol{e}_{\rm max}^{(i)} == q_{\rm value}$}
	% 			    \STATE $\boldsymbol{\hat{q}}_{\rm max}^{(f)} = i$
	% 			\ENDIF
 %        		\ENDALOOP
	%     		\STATE $\boldsymbol{e}_{\rm max} \rightarrow \boldsymbol{\hat{q}}_{\rm max}^{(f)}$
 %   		\ENDALOOP
	% \end{algorithmic}
\end{algorithm}

@@

\subsection{Eigen Similarity Analysis}
\label{sec:2_prop_EigenSimilarityAnalysis}

@@

This subsection describes the proposed eigen similarity analysis for detailed attack identification, in complement to the attack estimation carried out through MOS schemes and eigenvalue analysis. In Subsection \ref{sec:2_prop_TimeSimilarityAnalysis} we present the eigen similarity analysis for identification of time under attack. Next, in Subsection \ref{sec:2_prop_PortSimilarityAnalysis}, we show how to apply the eigen similarity analysis in order to identify network ports under attack.

\subsubsection{Time Similarity Analysis}
\label{sec:2_prop_TimeSimilarityAnalysis}

@@

Figure \ref{fig:2_fig8} ilustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\boldsymbol{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected.

\begin{figure}[h!]
     \includegraphics[width=15cm]{figures/incremental.eps}
     \caption{Traffic selection for incremental approach.}
     \label{fig:2_fig8}
\end{figure}

@@

The eigen similarity analysis can also be applied in an individual fashion, where each $n$-th minute must be individually appended into $\boldsymbol{X}^{(q)}$, as shown by Figure \ref{fig:2_fig9}.

\begin{figure}[h!]
     \includegraphics[width=15cm]{figures/individualized.eps}
     \caption{Traffic selection for individual approach.}
     \label{fig:2_fig9}
\end{figure}

The incremental and the individual approaches can be combined to obtain the incremental individualized approach, where each minute is incrementally appended into the selected $\boldsymbol{X}^{(q)}$ for obtaining $\boldsymbol{v}_{(n)}$ to similarity analysis of the $n$-th minute, until detect the first $n$-th minute under attack. Subsequently, $\boldsymbol{X}_n$ becomes the new reference of traffic without network attack and each subsequent minute must have its similarity individually evaluated, as shown in Figure \ref{fig:2_fig2}.

\begin{figure}[h!]
     \includegraphics[width=15cm]{figures/incremental_individualized.eps}
     \caption{Traffic selection for incremental individualized approach.}
     \label{fig:2_fig2}
\end{figure}

This approach of incremental similarity analysis followed by individual analysis after an attack detection allows to identify the attack period, highlighting the first and last time under attack. This identification is possible due to the variation of the most significant eigenvectors, which becomes more significant when compared a traffic under attack against a traffic with no attack, according to results which are discussed in Section \ref{sec:2_experimentalresults}.

\subsubsection{Port Similarity Analysis}
\label{sec:2_prop_PortSimilarityAnalysis}

@@

\section{Experiments and Results}
\label{sec:2_experimentalresults}

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:2_AnalyzedScenario}, the experimental scenario adopted in the experiments is summarized. Then, Section \ref{sec:2_largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames for the experimental scenario. In Section \ref{sec:2_MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection in the simulated dataset. Section \ref{sec:2_EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:2_EigenSimilarityAnalysis} shows the results of similarity analysis for detailed flood and port scan identification for the experimental scenario. Section \ref{sec:2_DarpaEvaluation} presents the results of the largest eigenvalue analysis, model order selection and the eigenvalue analysis for flood and probe attack detection in the DARPA 1998 dataset.

\subsection{Experimental Scenario}
\label{sec:2_AnalyzedScenario}

@@

\subsection{Largest Eigenvalues Analysis}
\label{sec:2_largesteigenvaluesanalysis}

For the evaluation of MOS Schemes accuracy for flood and port scan detection, the framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:2_prop_getv}. Through eigenvalue analysis of traffic with flood or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/eigenvalues_synflood.eps} 
     \caption{Eigenvalues of the sample covariance matrix (synflood).}
     \label{fig:2_fig10}
\end{figure}

Figure \ref{fig:2_fig10} depicts the eigenvalues calculated from sample covariance matrix of the network traffic used to evaluate the synflood attack identification. In Figure \ref{fig:2_fig10}, the largest eigenvalue related to the simulated synflood attack ($q = 4$) stands out significantly from the other eigenvalues.

Figure \ref{fig:2_fig11} illustrates the eigenvalues calculated from sample covariance matrix of the matrix used for fraggle attack detection. In Figure \ref{fig:2_fig10}, the largest eigenvalue related to the simulated synflood attack ($q = 5$) stands out significantly from the other eigenvalues, in accordance with the result shown in Figure \ref{fig:2_fig10} for the synflood attack analysis.

\begin{figure}[h!]
	\centering
     \includegraphics[width=11cm]{figures/eigenvalues_fraggle.eps}
     \caption{Eigenvalues of the sample covariance matrix (fraggle).}
     \label{fig:2_fig11}
\end{figure}

Figure \ref{fig:2_fig12} depicts the eigenvalues calculated from covariance matrix of zero mean and unitary standard deviation variables, of the network traffic matrix evaluated for port scan detection.
